---
title: "EP11_practical_hw"
author: "Vincent Fischer"
date: '2022-05-23'
output: html_document
---

---
title: "Using long-range interactions"
author: "Pierre-Luc"
date: "5/20/2022"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(AnnotationHub)
  library(ensembldb)
  library(InteractionSet)
  library(epiwraps)
})
```

## Obtain the data

```{r}
# p300 peaks in A549 cells:
download.file("https://www.encodeproject.org/files/ENCFF776TFX/@@download/ENCFF776TFX.bed.gz", "p300.narrowPeak.gz")
# Hi-C long-range interactions in A549 cells:
download.file("https://www.encodeproject.org/files/ENCFF803ZOW/@@download/ENCFF803ZOW.tsv", "interactions.tsv")
```

Load the interaction data, and form an GInteractions object:

```{r}
hic <- read.delim("interactions.tsv")
anchor1 <- GRanges(hic$chr1, IRanges(hic$x1, hic$x2))
anchor2 <- GRanges(hic$chr2, IRanges(hic$y1, hic$y2))
gi <- GInteractions(anchor1, anchor2, observed=hic$observed)
gi
```

Load the peaks:

```{r}
peaks <- rtracklayer::import("p300.narrowPeak.gz")
seqlevelsStyle(peaks) <- "ensembl"
```

## Link genomic interactions to TSS

We'll need this function:

```{r}
gi2targets <- function(gi, proms){
  stopifnot(is(gi, "GInteractions"))
  if(is(proms, "GRanges")){
    stopifnot(!is.null(proms$gene_name))
    seqlevelsStyle(proms) <- "ensembl"
    proms <- proms[proms$gene_name!=""]
  }else if(is(proms, "EnsDb")){
    # obtain the promoters
    proms <- trim(suppressWarnings(promoters(
      transcripts(ensdb, columns=c("gene_name")),
      upstream=100,downstream=50)))
    proms <- proms[proms$gene_name!=""]
  }else{
    stop("Unrecognized `proms` format")
  }
  seqlevelsStyle(gi) <- "ensembl"
  # find overlaps between first anchor and promoters
  o <- as.data.frame(findOverlaps(first(gi), proms))
  o$gene <- proms$gene_name[o$subjectHits]
  o <- o[!duplicated(o[,c(1,3)]),]
  o1 <- splitAsList(o$gene, o$queryHits)
  # find overlaps between second anchor and promoters
  o <- as.data.frame(findOverlaps(second(gi), proms))
  o$gene <- proms$gene_name[o$subjectHits]
  o <- o[!duplicated(o[,c(1,3)]),]
  o2 <- splitAsList(o$gene, o$queryHits)
  # merge anchors with an overlap in the mate anchor
  hic2 <- c( first(gi)[as.numeric(names(o2))], second(gi)[as.numeric(names(o1))] )
  # add target gene
  hic2$gene <- c(o2,o1)
  sort(hic2)
}
```

We first obtain the gene annotation (takes a long time if you don't already have the annotation):

```{r}
ah <- AnnotationHub()
ensdb <- ah[["AH95744"]] # human ensembl 104 annotation
```

As a shortcut, if you don't want to wait for the annotation to download, you can download and use these promoters:

```{r, eval=FALSE}
download.file("https://ethz-ins.org/content/w11_promoters.rds", "promoters.rds")
proms <- readRDS("promoters.rds")
```


We then use the function to link distal regions to the gene(s) whose TSS they are in contact with:

```{r}
gi2 <- gi2targets(gi,ensdb)  # or gi2 <- gi2targets(gi,proms)
gi2
```

## Checking distal targets of peaks

```{r}
# find overlaps
o <- findOverlaps(peaks, gi2)
# add target genes to the peaks
mcols(peaks)[from(o),"target"] <- gi2[to(o)]$gene
mcols(peaks)$target <- CharacterList(mcols(peaks)$target)
# look at the output
peaks
```
### Working with AtomicLists

Since one distal region can be in contact with multiple targets, the target column can contain multiple genes. In this case, we encode this using a `CharacterList` object, an extension of the `AtomicList` object, which is a powerful way of dealing with lists of atomic elements in R.

Let's take the target field for the first 10 peaks to play with a bit:

```{r, eval=FALSE}
cl <- head(peaks$target,10)
cl
lengths(cl)
cl=="ITGB1"
any(cl=="ITGB1")
```
## Homework

```{r}
# annotation of distal regions linked to TSS  
anno <- annotateRegions(peaks, anno = ensdb, proximal = c(2500,10000))
#subsetting for distance proximal to  TSS for 2.5kb to 10kb
#anno$class
anno2.5_10 <- anno[anno$class =="proximal >2500&<=10000bp"]
#anno2.5_10
# sum of interactions in proximal regions of tss by p300
interact_2.5_10 <-sum(any(!is.na(anno2.5_10$target)))
interact_2.5_10
# sum of interactions in proximal regions of tss by p300 where interactions == nearest TSS
targ_gene_na <-anno2.5_10$target==anno2.5_10$nearestTSS.gene_name # including NA targets
target_gene_nona<- sum(na.omit(any(targ_gene_na))) #excluding NA targets
target_gene_nona
perc_2.5_10 <- (target_gene_nona/interact_2.5_10)*100
perc_2.5_10
```
Of the genes that are between 2.5 and 10kb from the nearest TSS, `r perc_2.5_10` % (14.36%) form an interaction with that nearest gene.

```{r}
# annotation of distal regions linked to TSS  
annoX <- annotateRegions(peaks, anno = ensdb, proximal = c(10000,3e+09)) #3e+09 for max bp length, whole genome
#subsetting for distance proximal to  TSS for  >10kb
#anoX$class
anno_bigger_10 <- annoX[annoX$class =="proximal >10000&<=3e+09bp"]
#anno_bigger_10
# sum of interactions in proximal regions of tss by p300
interact_bigger_10 <-sum(any(!is.na(anno_bigger_10$target)))
interact_bigger_10
# sum of interactions in proximal regions of tss by p300 where interactions == nearest TSS
targ_gene_bigger_na <-anno_bigger_10$target==anno_bigger_10$nearestTSS.gene_name # including NA targets
target_gene_bigger_nona<- sum(na.omit(any(targ_gene_bigger_na))) #excluding NA targets
target_gene_bigger_nona
perc_bigger_10 <- (target_gene_bigger_nona/interact_bigger_10)*100
perc_bigger_10

```
Of the genes that are more than 10kb away from the nearest TSS, `r perc_bigger_10` % (20.36%) form an interaction with that nearest gene. Relative more distant p300 interactions (>10kb)happen than in close proximity.

```{r}
sessionInfo()
```

