---
title: "Epigenome2_Homework"
author: "Vincent Fischer"
date: "7 3 2022"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
})
```

```{r}
ah <- AnnotationHub()
#display(ah)
```
#load annotation GRCm38 version 102

```{r}
ensdb_m102 <- ah[["AH89211"]]
```
#load genome sequence 2bit GRCm38 GRCm38.p6	2020-10-27
```{r}
ensdb_m2bit <- ah[["AH88477"]]
ensdb_m2bit
```
#load genome sequence 2bit BDGP6 (Drosophila melanogaster) BDGP6.32 2021-10-20
```{r}
ensdb_d2bit <- ah[["AH99424"]]
```
#How many different ensembl gene IDs and gene symbols are there for protein-coding genes?
#Protein coding transcripts
```{r}
#ensdb_m102
txm <- transcripts(ensdb_m102)
codingm <- txm$tx_biotype == "protein_coding"
#codingm
codingnumbertx <- sum(codingm)
codingnumbertx
#protin coding transcrpits are 60320

```
#Protein coding genes
```{r}
#ensdb_m102
gm <- genes(ensdb_m102)
#gm
codinggm <- gm$gene_biotype == "protein_coding"
#codinggm
codingnumbergm <- sum(codinggm)
codingnumbergm
#protin coding genes are 22287
```
#Plot the distribution (histogram) of how many exons protein-coding genes
```{r}
#transcripts
exsts <- exons(ensdb_m102, column=c("tx_id"))
exsPerTx <- exonsBy(ensdb_m102, column=c("tx_id","tx_biotype"),
                  filter=TxBiotypeFilter("protein_coding"))
nbExonsPerPCtx <- lengths(exsPerTx)
hist(nbExonsPerPCtx)

#genes
exsPerGene <- exonsBy(ensdb_m102, column=c("gene_id","gene_biotype"),
                  filter=GeneBiotypeFilter("protein_coding"))
#exsPerGene
nbExonsPerPCGene <- lengths(exsPerGene)
hist(nbExonsPerPCGene)


```

#Plot the distribution of the (spliced) length of protein-coding transcripts
```{r}
#as.numeric(unlist(width(exsPerTx)))
head(width(exsPerTx))
hist(as.numeric(unlist(width(exsPerTx))), xlim = c(0,6000))
hist(mean(width(exsPerTx)), xlim = c(0,5000))
##stranges histograms what went wrong ?
#breaks as in number of cells number of width(exsPerTx) function
hist((as.numeric(unlist(width(exsPerTx)))),breaks = 541340, xlim = c(0,5000))

?hist
```
```{r}
sessionInfo()
```
# around 2 hours
